Online Multiclass Linear Classification with Bandit Feedback

### 2019-01-17 ICML-submission/algorithm-for-strongly-linearly-separable-data.tex

In this section, we consider the case when the examples are strongly linearly
separable. We present an algorithm for this setting
(Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples})
and we give an upper bound on its number of mistakes, stated as
\autoref{theorem:strongly-separable-examples-mistake-upper-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples} in the
supplementary material.

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm; see e.g.
\citet[Section 3.3.1]{Shalev-Shwartz-2012}. Each copy corresponds to one of the
classes. At each timestep $t$, upon seeing example $x_t$, copy $i$ makes a
binary prediction on whether $x_t$ belongs to class $i$ or not. If at least one
copy makes a positive prediction, the algorithm chooses its prediction as one of
the classes corresponding to a positive prediction. If all copies make negative
predictions, the algorithms makes a prediction uniformly at random.

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-examples-mistake-lower-bound}
Let $\gamma$ be a positive real number, and $R$ be a non-negative real
number. For any (possibly randomized) algorithm $\calA$ for the \textsc{Online
Multiclass Classification with Bandit Feedback} with $K \le
(R/\gamma)^2$ classes there exists an inner product space $(V,
\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ in $V \times
\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
number of mistakes $\calA$ makes is at least $\frac{K-1}{2} \left\lfloor
\frac{1}{4} (R/\gamma)^2 \right\rfloor$.
\end{theorem}

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $k(\cdot, \cdot)$ be a positive definite kernel function and let
$\phi(\cdot)$ be the associated feature map. Let $K$ be a positive integer,
$\gamma$ be a positive real number, $R$ be a non-negative real number. If $L =
((x_1, y_1), (x_2, y_2), \dots, (x_T, y_T))$ is a sequence of labeled examples
such that:
\begin{enumerate}
  \item $((\phi(x_1), y_1), (\phi(x_2), y_2), \dots, (\phi(x_T), y_T))$ is strongly linearly separable with margin $\gamma$,
  \item for all $t$ in $\cbr{1,\ldots,T}$, $k(x_t,x_t) \le R^2$,
\end{enumerate}
then the expected number of mistakes Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

>>>>

In this section, we consider the case when the examples are strongly linearly
separable. We present an algorithm for this setting
(Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples})
and give an upper bound on its number of mistakes, stated as
\autoref{theorem:strongly-separable-examples-mistake-upper-bound} below. The
proof of the theorem can be found in
Appendix~\ref{section:proofs-for-stringly-separable-examples} in the
supplementary material.

The idea behind
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples} is
to use $K$ copies of the \textsc{Binary Perceptron} algorithm; see e.g.
\citet[Section 3.3.1]{Shalev-Shwartz-2012}. Each copy corresponds to one of the
classes. At each timestep $t$, upon seeing example $x_t$, copy $i$ makes a
binary prediction on whether $x_t$ belongs to class $i$ or not. If at least one
copy makes a positive prediction, the algorithm chooses its prediction as one of
the classes corresponding to a positive prediction. If all copies make negative
predictions, the algorithms makes a prediction uniformly at random from
$\cbr{1,\ldots,K}$.

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-examples-mistake-lower-bound}
Let $\gamma$ be a positive real number, and $R$ be a non-negative real
number. For any (possibly randomized) algorithm $\calA$ for the \textsc{Online
Multiclass Classification with Bandit Feedback} problem with $K \le
(R/\gamma)^2$ classes, there exists an inner product space $(V,
\ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ in $V \times
\{1,2,\dots,K\}$ that are strongly linearly separable with margin $\gamma$, the
norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the expected
number of mistakes $\calA$ makes is at least $\frac{K-1}{2} \left\lfloor
\frac{1}{4} (R/\gamma)^2 \right\rfloor$.
\end{theorem}

@@

\begin{theorem}[Mistake upper bound for kernelized algorithm]
\label{theorem:kernelized-upper-bound}
Let $k(\cdot, \cdot)$ be a positive definite kernel function and let
$\phi(\cdot)$ be the associated feature map. Let $K$ be a positive integer,
$\gamma$ be a positive real number, $R$ be a non-negative real number. If $L =
((x_1, y_1), (x_2, y_2), \dots, (x_T, y_T))$ is a sequence of labeled examples
such that:
\begin{enumerate}
  \item $((\phi(x_1), y_1), (\phi(x_2), y_2), \dots, (\phi(x_T), y_T))$,
  the set of examples in $L$ after feature transformation $\phi$,
  is strongly linearly separable with margin $\gamma$,
  \item for all $t$ in $\cbr{1,\ldots,T}$, $k(x_t,x_t) \le R^2$,
\end{enumerate}
then the expected number of mistakes Algorithm~\ref{algorithm:kernelized} makes
is at most $(K-1) \lfloor 4(R/\gamma)^2 \rfloor$.
\end{theorem}

### 2019-01-17 ICML-submission/appendix-proofs-of-polynomial-approximation-theorems.tex

@@

@@

For convenience we define univariate rational function
\begin{align*}
S_{n,k}(z) & = \frac{A_{n,k}(z)}{B_{n,k}(z)} && \text{for $n,k \ge 0$,}
\end{align*}
and a multivariate rational function
$$
Q(x) = \left( \sum_{i=1}^m S_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \right) - \left(m - \frac{1}{2} \right) \; .
$$
It easy to verify that
$$
p(x) = Q(x) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
$$

>>>>

In this section, we follow the construction of~\citet{Klivans-Servedio-2008}
(which in turn uses the constructions of~\citet{Beigel-Reingold-Spielman-1995})
to establish two polynomials of low norm, such that it takes large positive values
in
\[ \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\} \]
and takes large negative values in
\[ \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le -\gamma \right\} . \]
We improve the norm bound analysis of~\citet{Klivans-Servedio-2008} in two aspects:
\begin{enumerate}
  \item Our upper bounds on the norm of the polynomials do not have any dependency on the
  dimensionality $d$.
  \item We remove the requirement that the fractional part of input $x$ must be above some threshold in
  Theorem~\ref{theorem:polynomial-approximation-2}.
\end{enumerate}
A lot of the proof details are similar to those of~\citet{Klivans-Servedio-2008}; nevertheless,
we provide a self-contained full proof here.

@@

We prove the following lemma in this section.
\autoref{theorem:polynomial-approximation-2} immediately
follows from this lemma by considering $p' = p \cdot 2^{-s(s+1)rm+1})$
and algebra.

\begin{lemma}
\label{lemma:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1},
\norm{v_2}, \dots, \norm{v_m} \le 1$. Let $\gamma \in (0,1)$.
Define
$$
r = 2 \left\lceil \frac{1}{4} \log_2(4m + 1) \right\rceil + 1 \quad \text{and} \quad s = \left \lceil \log_2(1/\gamma) \right \rceil \; .
$$
Then, there exists a multivariate polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $\displaystyle p(x) \ge \frac{1}{4} \cdot 2^{s(s+1)rm}$ \\
for all $\displaystyle x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$,

\item $\displaystyle p(x) \le - \frac{1}{4} \cdot 2^{s(s+1)rm}$ \\
for all $\displaystyle x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$,

\item $\deg(p) \le (2s+1) rm$,
\item $\norm{p} \le (2m-1/2) 2^m \cdot \left(2^{2s} rm (4s+2)^2 \right)^{(s+1/2)rm}$.
\end{enumerate}
\end{lemma}

@@

For convenience we define univariate rational function
\begin{align*}
S_{n,k}(z) & = \frac{A_{n,k}(z)}{B_{n,k}(z)} && \text{for $n,k \ge 0$,}
\end{align*}
and a multivariate rational function
$$
Q(x) = \left( \sum_{i=1}^m S_{s,r}\left( \frac{\ip{v_i}{x}}{\gamma} \right) \right) - \left(m - \frac{1}{2} \right) \; .
$$
It is easy to verify that
$$
p(x) = Q(x) \prod_{j=1}^m B_{s,r} \left( \frac{\ip{v_j}{x}}{\gamma} \right) \; .
$$

### 2019-01-17 ICML-submission/introduction.tex

The bandit feedback setting, however, is much more challenging. For the case
when the examples are strongly linearly separable, to the best of our knowledge,
it is not known how to design an efficient algorithm with a finite mistake bound
before our work.~\footnote{Although~\cite{Chen-Chen-Zhang-Chen-Zhang-2009}
claimed that their Conservative OVA algorithm with PA-I update has a finite
mistake bound under the strong linear separability condition, their proof of
Theorem 2 is incorrect. Their Lemma 4 cannot be directly applied to the bandit
setting, as the algorithm only makes updates on a subset of the $w_i$'s.} We
design a simple and efficient algorithm, namely
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
that has an expected number of mistakes at most $O(K (R/\gamma)^2))$.
The memory complexity and per-round time complexity are both $O(dK)$.
The algorithm can be viewed as running $K$ copies of the \textsc{Binary
Perceptron} algorithm, one copy for each class. Intuitively speaking, the $O(K)$
multiplicative factor in the mistake bound is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We prove
that any (possibly randomized) algorithm must make $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.
%$(K-1) \lfloor 4(R/\gamma)^2\rfloor$

>>>>

The bandit feedback setting, however, is much more challenging. For the case
when the examples are strongly linearly separable, to the best of our knowledge,
it is not known how to design an efficient algorithm with a finite mistake bound
before our work.~\footnote{Although~\cite{Chen-Chen-Zhang-Chen-Zhang-2009}
claimed that their Conservative OVA algorithm with PA-I update has a finite
mistake bound under the strong linear separability condition, their proof of
Theorem 2 is incorrect: their Lemma 4 cannot be directly applied to the bandit
setting, as the algorithm only makes updates on a subset of the $w_i$'s.} We
design a simple and efficient algorithm, namely
Algorithm~\ref{algorithm:algorithm-for-strongly-linearly-separable-examples},
that has an expected number of mistakes at most $O(K (R/\gamma)^2))$.
The memory complexity and per-round time complexity are both $O(dK)$.
The algorithm can be viewed as running $K$ copies of the \textsc{Binary
Perceptron} algorithm, one copy for each class. Intuitively speaking, the $O(K)$
multiplicative factor in the mistake bound is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We prove
that any (possibly randomized) algorithm must make $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.
%$(K-1) \lfloor 4(R/\gamma)^2\rfloor$

### 2019-01-17 ICML-submission/notions-of-linear-separability.tex

The notion of weak linear separability with a margin is standard. It is used in
the full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. It is a folklore result that if a set
of labeled examples is separable with a margin $\gamma$ and the norm of the
examples is bounded by $R$, then \textsc{Multiclass Perceptron} algorithm makes
at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore
result is that \textsc{Multiclass Perceptron} is essentially optimal in the
sense that any (possibly randomized) algorithm must make $\frac{1}{2}
\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we state the \textsc{Multiclass Perceptron} algorithm and give
proofs of both of these results in
Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
material.

\chicheng{Shall we mention that this multiclass margin notion has also appeared in e.g.~\cite{Crammer-Singer-2003}?}

>>>>

The notion of weak linear separability with a margin is standard;
using the notion of multiclass margin~\citep[see e.g.][]{Crammer-Singer-2003},
weak linear separability with margin $\gamma$ is equivalent to that
there is a linear classifier that has multiclass margin $\gamma$ on
all the examples. It is used in
the full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. It is a folklore result that if a set
of labeled examples is separable with a margin $\gamma$ and the norm of the
examples is bounded by $R$, then \textsc{Multiclass Perceptron} algorithm makes
at most $\left\lfloor 2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore
result is that \textsc{Multiclass Perceptron} is essentially optimal in the
sense that any (possibly randomized) algorithm must make $\frac{1}{2}
\left\lfloor (R/\gamma)^2 \right \rfloor$ mistakes in the worst case. For
completeness, we state the \textsc{Multiclass Perceptron} algorithm and give
proofs of both of these results in
Appendix~\ref{section:multiclass-perceptron-proofs} in the supplementary
material.

%\chicheng{Shall we mention that this multiclass margin notion has also appeared in e.g.?}
